A model-based method for measuring stability in personal culture with panel data[^thanks]

### 2020-12-21 ms_anonymized.Rmd
figure 2 updates and changing greek letters

\begin{eqnarray}
  y_{i1} = \alpha_1 + U_i + v_{i1} \\ 
  y_{i2} = \alpha_2 + U_i + v_{i2} \notag \\
  y_{i3} = \alpha_3 + U_i + v_{i3} \notag
\end{eqnarray}

From this basic version, we can add further constraints. For example we can constrain $\alpha_1 = \alpha_2 = \alpha_3$ if we are willing to assume that the mean response is the same at all waves. And we could constrain $\sigma^2_{v1} = \sigma^2_{v2} = \sigma^2_{v3}$ if we are willing to assume that the variance of the responses does not change over time.

This is not the place for a general overview of structural equation modeling [@bollen1989; see @kline2015 for an accessible introduction]. But the basic intuition in the three-wave case is that we are attempting to reconstruct 9 observed elements from the data (3 means, 3 variances, and 3 covariances) using a simpler model with fewer than 9 parameters. For example, if we make the assumptions in Equation (2), including the equality constraints in the following paragraph, we can estimate a model that has only 3 parameters: one value of $\alpha$ (shared by all waves), one $\sigma^2_v$ (shared by all waves), and one $\sigma^2_U$. Here $U$ is a "latent variable," which, in this specification, simply means a person-specific "error" (or fixed effect) that gets added equally to a person's response at every wave. $U_i$ reflects an individual's unchanging tendency to respond to a certain question the same way over the study period (a "settled disposition").

@@

\begin{eqnarray}
  y_{i3} = \alpha_3 + \rho y_{i2} + v_{i3} \\ 
  y_{i2} = \alpha_2 + \rho y_{i1} + v_{i2} \notag
\end{eqnarray}

Here we cannot make $Y_1$ a dependent variable because we don't have its previous value (which would be $Y_0$). Therefore we can only use $Y_1$ as a predictor. As above, we can make further simplifications if we want (e.g., $\alpha_2 = \alpha_3$ or $\sigma^2_{v2} = \sigma^2_{v3}$). Using this fully simplified version, we are estimating a model with 5 parameters: one value of $\alpha$ (shared by waves 2 and 3), one $\sigma^2_v$ (shared by waves 2 and 3), the $\rho$ autocorrelation parameter, and the mean and variance of $Y_1$.

@@

\begin{eqnarray}
  y_{i3} = \alpha_3 + \rho y_{i2} + U_i + v_{i3} \\ 
  y_{i2} = \alpha_2 + \rho y_{i1} + U_i + v_{i2} \notag \\
  \text{Cov}(U,Y_1) = \tau \notag
\end{eqnarray}

@@

Almost all (`r ok_fits[2,2] %>% round(.,1)` percent) of the preferred SDM models have acceptable fits (RMSEA < .08). But we can put the SDM to an even stronger test by estimating a model that assumes no updating across waves _and_ no changes in the means (i.e., no period effects). This is the model implied by Equation (2) and is equivalent to a simple confirmatory factor analysis with just three parameters: one value of $\alpha$ (shared by all waves), one $\sigma^2_v$ (shared by all waves), and one $\sigma^2_U$. In essence, this model assumes that each wave is a report of a fixed quantity determined before the time of the study.

>>>>

\begin{eqnarray}
  y_{i1} = \alpha_1 + U_i + \epsilon_{i1} \\ 
  y_{i2} = \alpha_2 + U_i + \epsilon_{i2} \notag \\
  y_{i3} = \alpha_3 + U_i + \epsilon_{i3} \notag
\end{eqnarray}

From this basic version, we can add further constraints. For example we can constrain $\alpha_1 = \alpha_2 = \alpha_3$ if we are willing to assume that the mean response is the same at all waves. And we could constrain $\sigma^2_{\epsilon_1} = \sigma^2_{\epsilon_2} = \sigma^2_{\epsilon_2}$ if we are willing to assume that the variance of the responses does not change over time.

This is not the place for a general overview of structural equation modeling [@bollen1989; see @kline2015 for an accessible introduction]. But the basic intuition in the three-wave case is that we are attempting to reconstruct 9 observed elements from the data (3 means, 3 variances, and 3 covariances) using a simpler model with fewer than 9 parameters. For example, if we make the assumptions in Equation (2), including the equality constraints in the following paragraph, we can estimate a model that has only 3 parameters: one value of $\alpha$ (shared by all waves), one $\sigma^2_{\epsilon}$ (shared by all waves), and one $\sigma^2_U$. Here $U$ is a "latent variable," which, in this specification, simply means a person-specific "error" (or fixed effect) that gets added equally to a person's response at every wave. $U_i$ reflects an individual's unchanging tendency to respond to a certain question the same way over the study period (a "settled disposition").

@@

\begin{eqnarray}
  y_{i3} = \alpha_3 + \rho y_{i2} + \epsilon_{i3} \\ 
  y_{i2} = \alpha_2 + \rho y_{i1} + \epsilon_{i2} \notag
\end{eqnarray}

Here we cannot make $Y_1$ a dependent variable because we don't have its previous value (which would be $Y_0$). Therefore we can only use $Y_1$ as a predictor. As above, we can make further simplifications if we want (e.g., $\alpha_2 = \alpha_3$ or $\sigma^2_{\epsilon_2} = \sigma^2_{\epsilon_3}$). Using this fully simplified version, we are estimating a model with 5 parameters: one value of $\alpha$ (shared by waves 2 and 3), one $\sigma^2_{\epsilon}$ (shared by waves 2 and 3), the $\rho$ autocorrelation parameter, and the mean and variance of $Y_1$.

@@

\begin{eqnarray}
  y_{i3} = \alpha_3 + \rho y_{i2} + U_i + \epsilon_{i3} \\ 
  y_{i2} = \alpha_2 + \rho y_{i1} + U_i + \epsilon_{i2} \notag \\
  \text{Cov}(U,Y_1) = \tau \notag
\end{eqnarray}

@@

Almost all (`r ok_fits[2,2] %>% round(.,1)` percent) of the preferred SDM models have acceptable fits (RMSEA < .08). But we can put the SDM to an even stronger test by estimating a model that assumes no updating across waves _and_ no changes in the means (i.e., no period effects). This is the model implied by Equation (2) and is equivalent to a simple confirmatory factor analysis with just three parameters: one value of $\alpha$ (shared by all waves), one $\sigma^2_{\epsilon}$ (shared by all waves), and one $\sigma^2_U$. In essence, this model assumes that each wave is a report of a fixed quantity determined before the time of the study.
